---
title       : Simple Linear Regression
subtitle    : Drawing straight lines through data
author      : Andrew Jackson
job         : Associate Professor
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, bootstrap, quiz]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

--- &two-cols w1:48% w2:48%

## Scatter plot data

*** {name: left}

> - How are two measures related?
> - Are they correlated?
> - Can one explain variation in the other?
> - What is the relationship?
> - Can I make new predictions using the relationship?

*** {name: right}

```{r, fig.width = 7, fig.height = 6, echo = FALSE}

# Author: Andrew Jackson
# Date: 08/10/2014
# About: An example R script that generates some random data according to a
# linear model and illustrates how you can gain insight into such relationships
# using the pairs() function, and also how you can interrogate the residuals
# of the corresponding fitted linear model for a posteriori insight.

library(viridis)
palette(viridis(8))

# Set the seed for the random number generator
set.seed(2)

# close all open graphics windows and remove all prior objects from the memory.
rm(list=ls())

# Create some randomly distributed x values representing predator/prey body mass
x <- runif(50,100,5000) / 1000

# generate some y values according to a linear model. Here, they have
# slope = 1, intercept = 1 and Gaussian (normal) random noise added with
# mean = 1 and standard deviation = 1. These y values represent the
# Interaction Strengths of the coupled predator-prey nodes in our 
# theoretical ecological community.
y <- 1.2*x + 1 + rnorm(50,0,1)

plot(x,y,
      ylab="Interaction Strength", xlab="Relative body mass (Predator/Prey)",
      las=1,bty="L",cex.lab=1.3,cex.axis=1.3, pch = 19, col = 5)

```

--- &two-cols w1:48% w2:48%

## Are they correlated?

*** {name: left}

> - Both x and y data show variation
> - Question is: do they co-vary (vary together)?
   - Are large $x$ values associated with large $y$ values?
   - or are large $x$ values associated with small $y$ values?
> - Calculate a statistics falle the correlation coefficient ($r$) which 
takes values $-1 \leq r \leq +1$
> - test $r$ against a statistical distribution to get a p-value using 
`cor.test()`

*** {name: right}

```{r, fig.width = 7, fig.height = 6, echo = FALSE}
# This function is used to put histograms on the diagonal
panel.hist <- function(x, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(usr[1:2], 0, 1.5) )
    h <- hist(x, plot = FALSE)
    breaks <- h$breaks; nB <- length(breaks)
    y <- h$counts; y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col=6, ...)
}

# This function is used to put (absolute) correlations on the upper panels,
# with size of the text proportional to the correlations.
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits=digits)[1]
    txt <- paste(prefix, txt, sep="")
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

# Here we create a dataframe object called "dat" which labels the x and y
# variables appropriately. The function pairs() needs the data as either a 
# dataframe or matrix type object.
dat <- data.frame(Relative.Body.Size = x, Interaction.Strength = y)


# create a new plotting window and call the pairs() function with the customised
# functions we defined above to make it more informative than the defaults. 
# The function panel.smooth() is in-built in R.
suppressWarnings(pairs(dat,
      upper.panel=panel.smooth,
      lower.panel=panel.cor,
      diag.panel=panel.hist,
      col.smooth = 2, pch = 19)
      )
```


--- 

## Let's ask a different question

> - Instead of...
> - Is there a relationship between $x$ and $y$
> - I want to know...
> - What _is_ the relationship between $x$ and $y$
> - Need to fit a line through the data and we can use this to describe what the 
relationship is, both verbally and mathematically

--- 

## A straight line
```{r, fig.width = 9, fig.height = 6, echo = FALSE}
# Add the null hypothesis line which is a horizontal line at the mean of
# all the data.    
plot(x, y,
      ylab="Interaction Strength", xlab="Relative body size(Predator/Prey)",
      las = 1, bty = "L", 
      cex.lab = 1.3, cex.axis = 1.3, col = 5, pch = 19)

m1 <- lm(y~x) # fit a linear model of y against x
abline(m1, col = 2, lwd = 2, lty = 1)

```

---

## A null hypothesis
```{r, fig.width = 9, fig.height = 6, echo = FALSE}
# Add the null hypothesis line which is a horizontal line at the mean of
# all the data.    
plot(x, y,
      ylab="Interaction Strength", xlab="Relative body size(Predator/Prey)",
      las = 1, bty = "L", 
      cex.lab = 1.3, cex.axis = 1.3, col = 5, pch = 19)

m1 <- lm(y~x) # fit a linear model of y against x
abline(m1, col = 2, lwd = 2, lty = 1)

abline(mean(y), 0, col = 7, lwd = 3, lty = 2)     
```

--- &two-cols w1:58% w2:38%

## Residuals

*** {name: left}

```{r, fig.width = 9, fig.height = 6, echo = FALSE}
# Add the null hypothesis line which is a horizontal line at the mean of
# all the data.    
plot(x, y,
      ylab="Interaction Strength", xlab="Relative body size(Predator/Prey)",
      las = 1, bty = "L", 
      cex.lab = 1.3, cex.axis = 1.3, col = 5, pch = 19)

m1 <- lm(y~x) # fit a linear model of y against x
abline(m1, col = 2, lwd = 2, lty = 1)

# extract the residuals from the linear model m1
rsd <- resid(m1)

# find the index address of the smallest (most negative) and largest residuals.
mn <- which.min(rsd)
mx <- which.max(rsd)

# add red lines connecting them to the corresponding estimated value from the 
# linear model which is extracted from the m1 object by calling fitted().
lines(c(x[mn],x[mn]), c(y[mn],fitted(m1)[mn]), col = 2, lwd = 2, lty = 2)
lines(c(x[mx],x[mx]), c(fitted(m1)[mx],y[mx]), col = 2, lwd = 2, lty = 2)

```

*** {name: right}

> - Computer fits the line by minimising the residuals off the line
>- Strictly (for least squares fitting), it minimises the sum of the squares
of the residuals
> - $\sum (y_i - \hat{y_i}) ^ 2$



---

## Residuals are informative
> - tells us which data are larger than predicted, and which are lower
> - Should ideally be normally distributed around the line
> - Test this with visual plots like histograms or q-q plots
> - Should be evenly spread around the line with no obvious trend

--- &two-cols w1:48% w2:48%

## Back to our line


*** {name: left}

> -  This figure shows how much algal growth occurs in one week in freshwater 
lakes of varying pollution levels
> -  It would be useful for us to know how much pollution affects algal growth
> -  It would also be useful if we could make predictions for growth rates in 
lakes we havent been able to observe such as:
> -  pristine unpolluted lakes
> -  or lake with exactly 2.2 mg/L pollution.

*** {name: right}

```{r, fig.width = 8, fig.height = 8, echo = FALSE}
# create some sample data for working on in lectures

# manually set the random number generator to start from same place each time
set.seed(100)

# ------------------------------------------------------------------------------
# create a line of the form y = b0 + (b1 * x)

# the intercept of the lines
b0 <- 50

# the slope of the lines
b1 <- -3

# some x data on which to base the line
x <- 1:11

# number of data points to create
n <- length(x)

# create some random data with slope b1 and standard deviation 3
y1 <- b0 + b1*x + rnorm(n,0,3)

# create some random data with slope b1 and standard deviation 10
y2 <- b0 + b1*x + rnorm(n,0,10)

# create some random data with slope 0 and standard deviation 3
y3 <- b0 + 0*x + rnorm(n,0,1)

# create some random data with slope 0 and standard deviation 10
y4 <- b0 + 0*x + rnorm(n,0,10)

# find the min and max y values across all the four lines so we can 
# plot them neatly below on the exact same scales
min.y <- min(c(y1,y2,y3,y4))
max.y <- max(c(y1,y2,y3,y4))

# use the glm() function to find the best fit line to each of the four
# sets of data. I will explain this in more detail later in the module.
model1 <- glm(y1~x)
model2 <- glm(y2~x)
model3 <- glm(y3~x)
model4 <- glm(y4~x)

plot(x, y2, 
     type="p", cex=1.3, lwd=2, 
     ylim=c(min.y-1,max.y+1),
     xlim = c(0, 12),
     xlab="Pollutant (mg/L)",
     ylab="growth (mm)", 
     bty = "L", las = 1,
     cex.lab = 1.3, cex.axis = 1.3,
     col = 5, pch = 19)

abline(model2,lwd=2,lty=2, col = 2)


```

--- 

## Find the equation of the line



