---
title: "Introduction to AIC-based Model Selection using polynomials"
author: "Andrew L Jackson"
date: "`r format(Sys.Date(), '%d-%b-%Y')`"
output: html_notebook
---

```{r, echo = FALSE}
# set some global knitr build options, notably the default figure size
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", 
                      fig.width = 6, fig.height = 5)
```

```{r, echo = TRUE}

# Start off with some house-keeping and set-up.

# load the viridis package for pretty and accessible colours especially for 
# those with colour blindness. You will need to install this if you dont have it 
# already using install.packages("viridis")
library(viridis)

## load (and install if required) these two packages if you want to 
## produce the ggplot version.
# library(ggplot2) 
# library(broom)

# Here I create a palette of 8 colours from the viridis spectrum. I can then 
# access these colours within plotting commands by e.g. plot(0,0, col = 1), 
# or plot(0,0, col = 4) or plot(0,0, col = 8). This palette is also useful as it
# will print clearly in greyscale without any extra effort. See vignette for 
# viridis for more details.
palette(viridis(8))

```

## Import and plot our data

We are going to work with a simple dataset that comprises the species richness 
of fishing hauls (trawl fishing) of varying effort (time in minutes). These 
data are contained in the file `curved-data.csv`.

```{r import-and-plot, echo = TRUE}
mydata <- read.csv("curved-data.csv", header = T, stringsAsFactors = FALSE)

plot(richness ~ effort, data = mydata, bty = "L", cex.lab = 1.2, 
     xlab = "Fishing Effort (time spent hauling in minutes)",
     ylab = "species richness")

# ## ----------------------------------------------------------------------
# ## alternatively plot with ggplot
# g_original <- ggplot(data = mydata, 
#                      mapping = aes(x = effort,
#                                    y = richness)) + 
#         geom_point()
# 
# print(g_original)
# ## ----------------------------------------------------------------------
```

The aim is to fit a set of polynomial models to the data and use AIC to identify 
the most parsimonious order polynomial. To get you started, we will first fit three 
models: the null model, a basic linear model, and a quadratic model. There after, 
you will fit higher order polynomials, compare their AIC and use the `predict` 
function to generate the esimated values for adding the fitted curves to the plot.


## The Null model (zero order polynomial)
The null model is the most simple linear model that we might fit to some data. 
It comprises only a mean and standard deviatoin (or variance depending on your 
preference) for the data. Mathematically it looks like $Y = \beta_0 + \epsilon$, 
where $\beta_0$ is the intercept and overall mean of the response variable $Y$ 
and $\epsilon$ is the error which is defined by a variance or standard 
deviation depending on your preference for notation.

The null model is specified by the function $Y \tilde 1$ where `1` represents the 
coefficient of the intercept which is always equal to `1` which is why we omit 
it from the equation $Y = \beta_0 * (1) + \epsilon$. We are going to get into 
the habit of using the `data = ` argument when we call `glm()`. This approach 
basically creates a private environment within the function glm() where it can 
see inside the dataframe object `mydata` and access the columns of data without 
having to manually use the dollar sign notation `mydata$richness`. This is 
important for a subsequent step where we will use the `predict` function.

**Note** that I am now using functions within the `plot()` command in order to 
make use of the `data =` argument that follows. This also has the benefit of 
being consistent with how we specify the models in `glm()`.

```{r, echo = TRUE}

null.model <- glm(richness ~ 1, data = mydata)

summary(null.model)


# the null model is just a mean, so graphically it is a horizontal line,
# and ordinarily we would not bother to plot it, as it is pretty much 
# utterly uniformative.
plot(richness ~ effort, 
     data = mydata, 
     bty = "L", 
     cex.lab = 1.2, 
     xlab = "Fishing Effort (time spent hauling in minutes)",
     ylab = "species richness",
     col = 1)

# use abline(h=) to add a horizontal line by extracting the intercept
# using coef() on the null.model object.
abline(h = coef(null.model) , col = 6, lty = 2, lwd = 2)

# histogram of the residuals...
hist(resid(null.model), 15, col = 7)


# ## ----------------------------------------------------------------------
# # ggplot alternative. We can simply take g_original and modify it
# g_null <- g_original + geom_abline(slope = 0,
#                                    intercept = coef(null.model), 
#                                    col = "green",
#                                    linetype = 2)
# print(g_null)
# ## ----------------------------------------------------------------------

```

It is important to note the AIC of this fitted model which is 
`r round(extractAIC(null.model)[2], digits = 1)`. There is no need to check the 
residuals of this fitted model beyond a simple histogram which shows that they 
are not normally distributed. Clearly the line does not explain the data well, 
and you can quickly determine that this is not a very good model of the data. 
This is hardly surprising though given how simple it is: on to more complicated 
models!

## Simple linear model (1st order polynomial)
We don't expect a straight line model to be much better, but its still worth 
fitting it for completeness and by way of providing a comparison against which 
we can test a quadratic model in the next step.

```{r, echo = TRUE}

first.model <- glm(richness ~ effort, data = mydata)

summary(first.model)


plot(richness ~ effort, 
     data = mydata, 
     bty = "L", 
     cex.lab = 1.2, 
     xlab = "Fishing Effort (time spent hauling in minutes)",
     ylab = "species richness",
     col = 1)

# use abline(h=) to add a horizontal line by extracting the intercept
# using coef() on the null.model object.
abline(first.model , col = 6, lty = 2, lwd = 2)

# create a 2-panel plotting window for the diagnostics on the 
# residuals
par(mfrow=c(1,2))
# histogram of the residuals...
hist(resid(first.model), 15, col = 7)

# plot of the residuals of the line as a function of effort
plot(resid(first.model) ~ effort, data = mydata)
abline(h = 0 , col = 6, lty = 2, lwd = 2)

# ## ----------------------------------------------------------------------
# ## ggplot alternative
# g_poly1 <- g_original + 
#         geom_abline(intercept = coef(first.model)[1],
#                     slope = coef(first.model)[2], 
#                     color = "green")
# print(g_poly1)

# ## ggplot using geom_sooth instead with a linear fit
# g_poly1_version2 <- g_original + 
#         geom_smooth(method = "lm")
# print(g_poly1_version2)
# ## ----------------------------------------------------------------------
```

Again, note the AIC of this fitted model which is 
`r round(extractAIC(first.model)[2], digits = 1)`. The residual plots are more 
helpful to us now, but still they are neither normally distributed, nor 
randomly scattered either side of the expected line, and show marked pattern 
with respect to `effort`. On to a second-order polynomial.

## A 2nd order polynomail linear model (a quadratic)

In order to include the square of effort in the model, the most appropriate way to do this is to include a new piece of syntax within the `glm()` function that takes the following generic form: `glm(y ~ x + I(x^2) )`. Note that we wrapped the x squared operation inside a `I()`. This is absolutely essential as it tells R and the glm function to interpret that as *do this calculate as-Is*. The reason this is essential as had we simply written x^2 then it would do something else entirely unintended (I wont go into what it does, but suffice to say it is not what we want). And the same would be true for higher polynomials e.g. `glm(y ~ x + I(x^2) + I(x^3))` etc...

There is another approach that involves creating new columns (vectors) of data in our dataset that are the various powers of the x variable. This is an ok approach, but it becomes really clunky later on. Using the `I(x^2)` notation is hugely powerful as when we later on ask R to predict values along a new sequence of `x` values, then R will automatically apply all the various polynomials included in our model. 


```{r, echo = TRUE}


# Include the square term of "effort" in the glm using the I() notation.
second.model <- glm(richness ~ effort + I(effort^2), data = mydata)

# print a summary of this model to check it is as we intended.
summary(second.model)


# You could of course plot this in ggplot if you prefer
par(mfrow = c(1,1)) # make sure we are back to plotting a single panel again
plot(richness ~ effort, 
     data = mydata, 
     bty = "L", 
     cex.lab = 1.2, 
     xlab = "Fishing Effort (time spent hauling in minutes)",
     ylab = "species richness",
     col = 1)

# Now are a bit stuck. We can no longer use abline to add the model 
# predictions as it only works with straight lines. We have 
# use the predict function, which will take a bunch of hypothetical 
# x values and calculate the corresponding y values based on the 
# estimated coefficients we got by fitting our model to our data.
# In order to do this, we need to create a set of dummy effort 
# values for which we want to calculate the corresponding estimated 
# richness values for plotting a nice curve.

# a sequence of hypothetical effort values from 0 to 100 in steps of 1.
# The smaller we make the step-size, the smoother the curve will appear.
new.effort <- seq(0, 100, 1)

# create a new data.frame object with columns that have the EXACT same 
# name as in our original data. We dont need the response value though, 
# since we are trying to estimate that using this method!
# Note that we dont need to include a column called effort^2 because 
# the model knows to expect "effort" and knows to square it because of our
# use of I(effort^2).
new.data <- data.frame(effort = new.effort)

# now predict the richness based on this new data which i call y.hat
# in reference to the mathematical notation of putting an arrow head 
# like a hat over predicted values. Here we predict using second.model,
# and pass it the hypothetical effort values to use to make the 
# predictions.
y.hat <- predict(second.model, newdata = new.data)

# and finally, add this line to the plot of the data as a function of 
# our hypothetical new.effort values
lines(y.hat ~ new.effort, data = mydata, col = 6, lty = 2, lwd =2)

# create a 2-panel plotting window for the diagnostics on the 
# residuals
par(mfrow=c(1,2))
# histogram of the residuals...
hist(resid(second.model), 15, col = 7)

# plot of the residuals of the line as a function of effort
plot(resid(second.model) ~ effort, data = mydata)
abline(h = 0 , col = 6, lty = 2, lwd = 2)

# ## ----------------------------------------------------------------------
# ## ggplot alternative. This gets a little more complicated now.
# ## Here we have to use the broom package to help with the prediction.
# ## It will "augment" our dataframe by adding some columns to it relating to the 
# ## model fit. We again start with our original basic plot but now add to it a 
# ## geom_line() using a new predicted dataset created with teh augment() function
# ## in the package broom which you may need to install. This then uses the model 
# ## object second.model to create predicted values using the new data we defined 
# ## in new.data. The key column here is now called ".fitted". In the mapping 
# ## geom_line() will inherit the aesthetic that maps effort as the x variable and 
# ## we overwrite the y variable mapping so that it is now associated with 
# ## our newly created ".fitted" variable.
# g_poly2 <- g_original + 
#         geom_line(data = broom::augment(second.model,
#                                  newdata = new.data),
#                   mapping = aes(y = .fitted),
#                   color = "green")
# 
# ## print our nice ggplot to screen
# print(g_poly2)
# ## ----------------------------------------------------------------------

```

Again, note the AIC of this fitted model which is 
`r round(extractAIC(second.model)[2], digits = 1)`. The residual plots are more 
helpful to us now. They are pretty well normally distributed and are on the 
whole closer to zero. But they are still not randomly scattered either side of 
the expected line, and show marked pattern with respect to `effort`. 

## Tasks: fit higher order polynomials
Your tasks for this week are as follows:

+ fit a series of higher order polynomials, from 3rd to 8th order. There is no 
need to plot the estimated models or check the residuals for orders 4 and above. 
But you may like to do this to see what the higher order polynomials look like. 
In fact, it is entirely possible to write a loop in R that would automate 
this... but I havent tried, and more often than not its a case of easier said 
than done, but by all means take up the challenge if you like and we can help 
you with it.
+ Compile a table of AIC values for all polynomials from 0 to 8 and make an 
argument for which polynomial best explains the data.


## Quickly fit the higher order polynomials

```{r}

poly0 <- glm(richness ~ 1, 
             data = mydata)

poly1 <- glm(richness ~ effort, 
             data = mydata)

poly2 <- glm(richness ~ effort + I(effort^2), 
             data = mydata)

poly3 <- glm(richness ~ effort + I(effort^2) + I(effort^3), 
             data = mydata)

poly4 <- glm(richness ~ effort + I(effort^2) + I(effort^3) + I(effort^4), 
             data = mydata)

poly5 <- glm(richness ~ effort + I(effort^2) + I(effort^3) + I(effort^4) + 
                     I(effort^5), 
             data = mydata)

poly6 <- glm(richness ~ effort + I(effort^2) + I(effort^3) + I(effort^4) + 
                     I(effort^5) + I(effort^6), 
             data = mydata)

poly7 <- glm(richness ~ effort + I(effort^2) + I(effort^3) + I(effort^4) + 
                     I(effort^5) + I(effort^6) + I(effort^7), 
             data = mydata)

poly8 <- glm(richness ~ effort + I(effort^2) + I(effort^3) + I(effort^4) + 
                     I(effort^5) + I(effort^6) + I(effort^7) + I(effort^8), 
             data = mydata)

```

Extract AIC for all

```{r}

AIC(poly0, poly1, poly2, poly3, poly4, poly5, poly6, poly7, poly8)

```

Polynomial order 5 is the most parsimonious based on AIC. Even though poly7 has the lowest AIC, it has 2 more parameters that poly5, and so would need to be more than $2 \times 2 = 4$ units lower, but it is only approximately 2.5 units lower.


## Check residuals

Assuming we did settle on poly5 as our selected model we would absolutely need to check the residuals before finalising our choice.

```{r}

# create a 2-panel plotting window for the diagnostics on the 
# residuals
par(mfrow=c(1,2))
# histogram of the residuals...
hist(resid(poly5), 15, col = 7)

# plot of the residuals of the line as a function of effort
plot(resid(poly5) ~ effort, data = mydata)
abline(h = 0 , col = 6, lty = 2, lwd = 2)



```




